Hippocampal segmentation with MAGeT

### 2013-01-24 paper.Rnw
Incorporated changes from some of Mallar's comments

\title{Hippocampal segmentation with MAGeT}
\author{Jon Pipitone}
\maketitle

\begin{abstract}
Neuroimaging research often relies on automated anatomical segmentations of MR
images of the brain. Current multi-atlas based approaches provide accurate
segmentations of brain images by propagating manually derived segmentations of
specific neuroanatomical structures to unlabelled data. These approaches often
rely on a large number of such manually segmented atlases that take significant
time and expertise to produce. We present an algorithm for the automatic
segmentation of the hippocampus that minimizes the number of atlases needed
while still achieving similar accuracy to multi-atlas approaches.
\todo[inline]{Finish this... }
\end{abstract}

@@

The hippocampus is of particular interest to many researchers because it is
implicated in forms of brain dysfunction such as Alzheimer's disease and
schizophrenia, and has functional significance in cognitive processes such as
learning and memory.  For many research questions involving magnetic resonance
imaging (MRI) data accurate identification of the hippocampus and its
subregions is a necessary first step to better understand the individual
neuroanatomy of subjects.

Currently, the gold standard for neuroanatomical segmentation is manual
delineation by an expert human rater.  This is problematic for segmentation of
the hippocampus for several reasons.  First, manual segmentation takes a
significant investment of time and expertise \citep{Hammers2003} which may not
be readily available to researchers or clinicians.  Second, the amount of data
produced in neuroimaging experiments increasingly exceeds the capacity for
identification of specific neuroanatomical structures by an expert manual
rater.  Third, the true definition of hippocampal anatomy in MR images is
disputed \citep{Geuze2004}, as evidenced by efforts to create an unified
segmentation protocol \citep{Jack2011}.

Compounding each of these problems is the significant neuroanatomical
variability in the hippocampus throughout the course of aging, development, and
neuropsychiatric disorders \citep{Mouiha2011b}.  The result is that existing
hippocampal atlases available to a researcher may not accurately represent
neuroanatomy of a specific population under study.  Additionally, in the course
of a research or clinical study, it may be necessary to make adjustments to
hippocampal definition as a means of hypothesis testing.  For example, Poppenk
\citep{Poppenk2011} found that subdividing the hippocampus into anterior and
posterior regions resulted in a predictive relationship between volume
difference of those regions and recollection memory performance.  Making
modifications to an set of atlas library segmentations may require the same
level of expertise and manual effort needed to create the segmentations from
scratch.

%  Automated methods 
%
%  Heckemann2006 - 
%  Heckemann2011 -  Automatic morphometry in AD and MCI 
%    - (MAPER). multi-atlas. 30 atlases, entire ADNI screening  
%  Collins2010 - .. seg of HC and AM by augmenting ANIMAL
%    - multi-atlas 80 atlases, atlas selection (NMI)
%  Aljabar2009 - atlas selection and effect on accuracy
%    - 
%  Leung2010 - X-sectional and longitudinal HC vol in MCI, AD
%    - (MAPS), multi-atlas + STAPLE. 55 atlases. baseline+1yr of ADNI seg.
%    - atrophy rates
%  Lotjonen2010 - Fast and robust
%    - IBSR/ADNI.  
%  Mouiha2011b - Hippocampal atrophy rates in AD
%    - compare SNT vs FreeSurfer on ADNI
%
Automated segmentation techniques overcome the need for human expertise by
performing segmentations computationally.  A popular class of automated
methods, {\it multi-atlas-based segmentation}, use a series of expertly
segmented MRIs (an atlas library) to act as templates against which a target
image is labelled
\cite{Aljabar2009,Collins2010,Heckemann2011,Leung2010,Lotjonen2010,Leung2010,Wolz2010}.
In the basic method, each atlas is warped to fit a subject's neuroanatomy using
nonlinear registration techniques \citep{Collins1995,Klein2009}.  Atlas labels
are then transformed by this warping and a {\it label fusion} technique, such
as voxel-wise voting, is used to merge the competing label definitions from
each atlas into a final segmentation for a subject.

Many descriptions of multi-atlas-based hippocampal segmentation algorithms
report relying on an atlas library containing anywhere between 30 and 80
expertly labeled images
\citep{Heckemann2011,Collins2010,Aljabar2009,Leung2010,Lotjonen2010} in order
to achieve performance approach that of a manual rater.  As already noted, the
production of an atlas library requires significant manual effort and
expertise, as does the modification of an existing atlas library in order to
better suit a target population or investigate a specific research hypothesis.
Therefore, in this paper we tackle the problem of producing accurate
segmentations using small numbers of manually segmented atlases.

>>>>

\title{Hippocampal segmentation with MAGeT}
\author{Pipitone J., Winterburn J., Lerch J., Pruessner J., Lepage M., Voineskos A., Chakravarty M.M., and the Alzheimer's Disease Neuroimaging Initiative}
\maketitle

\begin{abstract}
Neuroimaging research often relies on automated anatomical segmentations of MR
images of the brain. Current multi-atlas based approaches provide accurate
segmentations of brain images by propagating manually derived segmentations of
specific neuroanatomical structures to unlabelled data. These approaches often
rely on a large number of such manually segmented atlases that take significant
time and expertise to produce. We present an algorithm for the automatic
segmentation of the hippocampus that minimizes the number of atlases needed
while still achieving similar accuracy to multi-atlas approaches.
\todo[inline]{finish}
\end{abstract}

@@

The hippocampus is of particular interest to many researchers because it is
implicated in forms of brain dysfunction such as Alzheimer's
disease\citep{Sabuncu2011} and schizophrenia\citep{Narr2004,Karnik-Henry2012},
and has functional significance in cognitive processes such as learning and
memory\citep{DenHeijer2012,Scoville2000}.  For many research questions
involving magnetic resonance imaging (MRI) data accurate identification of the
hippocampus and its subregions is a necessary first step to better understand
the individual neuroanatomy of subjects.

Currently, the gold standard for neuroanatomical segmentation is manual
delineation by an expert human rater.  This is problematic for hippocampal
segmentation for several reasons.  First, manual segmentation takes a
significant investment of time and expertise \citep{Hammers2003} which may not
be readily available to researchers or clinicians.  Second, the amount of data
produced in neuroimaging experiments increasingly exceeds the capacity for
identification of specific neuroanatomical structures by an expert manual
rater.  Third, the true definition of hippocampal anatomy in MR images is
disputed \citep{Geuze2004}, as evidenced by efforts to create an unified
segmentation protocol \citep{Jack2011}.

Compounding each of these problems is the significant neuroanatomical
variability in the hippocampus throughout the course of aging, development, and
neuropsychiatric disorders \citep{Mouiha2011b,Sabuncu2011}.  The result is that
existing hippocampal atlases available to a researcher may not accurately
represent neuroanatomy of a specific population under study.  Additionally, in
the course of a research or clinical study, it may be necessary to make
adjustments to hippocampal definition as a means of hypothesis testing.  For
example, Poppenk \citep{Poppenk2011} found that subdividing the hippocampus
into anterior and posterior regions resulted in a predictive relationship
between volume difference of those regions and recollection memory performance.
Making modifications to an set of atlas library segmentations may require the
same level of expertise and manual effort needed to create the segmentations
from scratch.

%  Automated methods 
%
%  Heckemann2006 - 
%  Heckemann2011 -  Automatic morphometry in AD and MCI 
%    - (MAPER). multi-atlas. 30 atlases, entire ADNI screening  
%  Collins2010 - .. seg of HC and AM by augmenting ANIMAL
%    - multi-atlas 80 atlases, atlas selection (NMI)
%  Aljabar2009 - atlas selection and effect on accuracy
%    - 
%  Leung2010 - X-sectional and longitudinal HC vol in MCI, AD
%    - (MAPS), multi-atlas + STAPLE. 55 atlases. baseline+1yr of ADNI seg.
%    - atrophy rates
%  Lotjonen2010 - Fast and robust
%    - IBSR/ADNI.  
%  Mouiha2011b - Hippocampal atrophy rates in AD
%    - compare SNT vs FreeSurfer on ADNI
%
Fully-automated segmentation techniques overcome the need for human expertise.
In particular, {\it atlas-based} approaches make use of one or more expertly
segmented MRIs (the atlas library) as templates to guide the segmentation of a
query subject's MRI
\cite{Aljabar2009,Collins2010,Heckemann2011,Leung2010,Lotjonen2010,Leung2010,Wolz2010}.
The basic strategy is as follows. Each atlas is warped to fit a subject's
neuroanatomy using nonlinear registration techniques
\citep{Collins1995,Klein2009}.  Atlas labels are then transformed by this
warping to the subject's space, and a {\it label fusion} technique, such as
voxel-wise voting, is used to merge the competing label definitions from each
atlas into a final segmentation for the subject.

Multi-atlas hippocampal segmentation algorithms achieve some of the best
automated segmentation accuracies to-date, and approach the performance of a
manual rater
\citep{Heckemann2011,Collins2010,Aljabar2009,Leung2010,Lotjonen2010}, but rely
on an atlas library containing anywhere between 30 and 80 expertly labeled
images to achieve this accuracy.  As already noted, the production of an atlas
library requires significant manual effort and expertise, as does the
modification of an existing atlas library in order to better suit a target
population or investigate a specific research hypothesis.  Therefore, in this
paper we tackle the problem of producing accurate segmentations using small
numbers of manually segmented atlases.
